An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-13 aistats2017.tex
More minor typos

Many state-of-the-art machine learning methods, and deep learning in particular,
are based on minibatch updates (such as SGD) to a model.  Minibatch updates
produce many improvements to the model for each pass over the dataset, and have
high sample efficiency.  In contrast, conventional M-H requires calculations
over the full dataset to produce a new sample.  Recent results
from~\citet{cutting_mh_2014} and~\citet{icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets of the full dataset.
The amount of data consumed for each test varies significantly from one
minibatch to the next. By contrast,~\citet{conf/uai/MaclaurinA14,TallData15}
perform exact tests but require a lower bound on parameter distribution across
its domain.  The amount of data reduction depends on the accuracy of this bound,
and such bounds are only available for relatively simple distributions.

@@

    \item We compare performance of our new test and prior approaches on several
    datasets. We demonstrate orders of magnitude improvements in efficiency
    (measured as data consumed per test), and that it does not suffer from
    long-tailed minibatch sizes.
\end{itemize}

@@

In the Metropolis-Hastings method~\citep{gilks1996markov,brooks2011handbook}, a
difficult-to-compute probability distribution $p(\theta)$ is sampled using a
Markov chain $\theta_1,\ldots,\theta_n$. The sample $\theta_{t+1}$ at time $t+1$
is generated using a candidate $\theta'$ from a (simpler) proposal distribution
$q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test
is usually a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
    \alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability
$\alpha(\theta_t,\theta')$, we accept $\theta'$ and set $\theta_{t+1} =
\theta'$, otherwise set $\theta_{t+1}=\theta_t$.  The test is often implemented
with an auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison
$u<\alpha(\theta_t,\theta')$; here, $\mathcal{U}(a,b)$ denotes the uniform
distribution on the interval $[a,b]$.  For simplicity, we drop the subscript $t$
for the current sample $\theta_t$ and denote it as $\theta$.

@@

Since minibatches contains randomly selected samples $x_i$, the values
$\Lambda_i$ are i.i.d. random variables\footnote{The analysis assumes sampling
with replacement although implementations on typical large datasets will
approximate this by sampling without replacement.}.  By the Central Limit
Theorem, we expect $\Lambda^*(\theta,\theta')$ to be approximately Gaussian. The
acceptance test then becomes a statistical test of the hypothesis that
$\Lambda(\theta,\theta')>\psi(u,\theta,\theta')$ by establishing that
$\Lambda^*(\theta,\theta')$ is substantially larger than
$\psi(u,\theta,\theta')$.

@@

\begin{lemma}\label{lem:worst_case}
    For the model in Section~\ref{ssec:gaussian_example}, there exists a fixed
    (independent of $N$) constant $c$ such that with probability $\geq c$ over
    the joint distribution of $(\theta, \theta', u)$, the tests
    from~\cite{cutting_mh_2014,icml2014c1_bardenet14} consume all $N$ samples. 
\end{lemma}
\vspace{-1em}
\begin{proof}
See Supplementary Material, Section~\ref{app:worst_case_proof}.
\end{proof}
Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta',u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated tests from prior work require at least a constant fraction $\geq c$
in the amount of data consumed per test compared to full-data tests, so their
speed-up is $\le 1/c$. The issue is the use of tail bounds to separate $\Delta$
from zero; for certain input/random $u$ combinations, $\Delta$ can be
arbitrarily close to zero. We avoid this by using the {\em approximately normal}
variation in $\Delta^*$ to {\em replace} the variation due to $u$.

>>>>

Many state-of-the-art machine learning methods, and deep learning in particular,
are based on minibatch updates (such as SGD) to a model.  Minibatch updates
produce many improvements to the model for each pass over the dataset, and have
high sample efficiency.  In contrast, conventional M-H requires calculations
over the full dataset to produce a new sample.  Recent results
from~\citet{cutting_mh_2014} and~\citet{icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets of the full dataset.
The amount of data consumed for each test varies significantly from one
minibatch to the next. By contrast,~\citet{conf/uai/MaclaurinA14,TallData15}
perform exact tests but require a lower bound on the parameter distribution across
its domain.  The amount of data reduction depends on the accuracy of this bound,
and such bounds are only available for relatively simple distributions.

@@

    \item We compare performance of our new test and prior approaches on several
    datasets. We show order-of-magnitude improvements in efficiency
    (measured as data consumed per test), and that it does not suffer from
    long-tailed minibatch sizes.
\end{itemize}

@@

In the Metropolis-Hastings method~\citep{gilks1996markov,brooks2011handbook}, a
difficult-to-compute probability distribution $p(\theta)$ is sampled using a
Markov chain $\theta_1,\ldots,\theta_T$. The sample $\theta_{t+1}$ at time $t+1$
is generated using a candidate $\theta'$ from a (simpler) proposal distribution
$q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test
is usually a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
    \alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability
$\alpha(\theta_t,\theta')$, we accept $\theta'$ and set $\theta_{t+1} =
\theta'$, otherwise set $\theta_{t+1}=\theta_t$.  The test is often implemented
with an auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison
$u<\alpha(\theta_t,\theta')$; here, $\mathcal{U}(a,b)$ denotes the uniform
distribution on the interval $[a,b]$.  For simplicity, we drop the subscript $t$
for the current sample $\theta_t$ and denote it as $\theta$.

@@

Since minibatches contains randomly selected samples $x_i$, the values
$\Lambda_i$ are i.i.d. random variables.\footnote{The analysis assumes sampling
with replacement although implementations on typical large datasets will
approximate this by sampling without replacement.} By the Central Limit
Theorem, we expect $\Lambda^*(\theta,\theta')$ to be approximately Gaussian. The
acceptance test then becomes a statistical test of the hypothesis that
$\Lambda(\theta,\theta')>\psi(u,\theta,\theta')$ by establishing that
$\Lambda^*(\theta,\theta')$ is substantially larger than
$\psi(u,\theta,\theta')$.

@@

\begin{lemma}\label{lem:worst_case}
    For the model in Section~\ref{ssec:gaussian_example}, there exists a fixed
    (independent of $N$) constant $c$ such that with probability $\geq c$ over
    the joint distribution of $(\theta, \theta', u)$, the tests
    from~\cite{cutting_mh_2014,icml2014c1_bardenet14} consume all $N$ samples. 
\end{lemma}
\vspace{-1em}
\begin{proof}
See Supplementary Material, Section~\ref{app:worst_case_proof}.
\end{proof}
Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta',u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated tests from prior work require at least a constant fraction $\geq c$
in the amount of data consumed per test compared to full-data tests, so their
speed-up is $\le 1/c$. The issue is the use of tail bounds to separate $\Lambda-\psi$
from zero; for certain input/random $u$ combinations, this difference can be
arbitrarily close to zero. We avoid this by using the {\em approximately normal}
variation in $\Lambda^*$ to {\em replace} the variation due to $u$.
